Solving Collective Spin Hamiltonians with Decoherence

### 2018-08-16 soc_squeezing/cumulant_expansions.tex
fixed some errors

Previous work has shown how to realize various spin-squeezing
Hamiltonians with ultracold atomic systems.  While it is
straightforward to simulate the dynamics generated by these
Hamiltonians in the context of a closed quantum system, here we try to
develop an understanding of their the squeezing behavior in the
presence of decoherence.  In particular, we will be considering
dynamics induced by the collective one-axis twisting (OAT), two-axis
twisting (TAT), and transverse field (TVF) Hamiltonians of the form
\begin{align}
  H_{\t{OAT}} = \chi S_z^2,
  &&
  H_{\t{TAT}} = \chi \p{S_z^2 - S_y^2},
  &&
  H_{\t{TVF}} = \chi S_z^2 - \Omega S_x,
\end{align}
for spins which decay in an uncorrelated manner at a rate $\gamma$.
The evolution of any Heisenburg operator $\O$ with a Hamiltonian $H$
is thus given by
\begin{align}
  \f{d}{dt} \O
  = i\sp{H,\O}_- + \gamma\D\p{\O},
  &&
  \D\p{\O} \equiv \sum_j\p{\sigma_j^+\O\sigma_j^-
    - \f12\sp{\sigma_j^+\sigma_j^-,\O}_+},
  \label{eq:EOM}
\end{align}
where $\sp{A,B}_\pm\equiv AB\pm BA$ and $j$ indexes an individual
spin.  Our measure of squeezing is
\begin{align}
  \xi^2 \equiv \f{N}{\abs{\bk{\v S}}^2}
  \min_{\uv n\perp\bk{\v S}} \bk{\p{\v S\c\uv n}^2},
  \label{eq:squeezing}
\end{align}
where $N$ is the total number of spins, $\v S$ is the collective spin
vector, and the minimization is performed over all unit vectors
$\uv n$ in the plane orthogonal to the mean spin vector $\bk{\v S}$.
This squeezing parameter is entirely determined by components
$\bk{S_\alpha}$ of the mean spin vector in addition to collective
spin-spin correlators of the form $\bk{S_\alpha S_\beta}$; our task is
therefore to compute these expectation values in all scenarios of
interest.

@@

This text frequently makes use of various spin operator identities; we
collect some particularly useful identities here for reference.
\begin{align}
  \f12 \sp{\sigma_j^z,\sigma_k^\mu}_- = \delta_{jk} \mu \sigma_j^\mu,
  &&
  \sp{S_z,\sigma_j^\mu}_- = \f12 \sp{\sigma_j^z,S_\mu}_-
  = \mu \sigma_j^\mu,
  &&
  \sp{S_z,S_\mu}_- = \mu S_\mu.
  \label{eq:comm_z_mu}
\end{align}
\begin{align}
  \sp{\sigma_j^\mu,\sigma_k^\nu}_-
  = \delta_{jk} \delta_{\mu,-\nu} \mu \sigma_j^z,
  &&
  \sp{S_\mu,\sigma_j^\nu}_- = \sp{\sigma_j^\mu,S_\nu}_-
  = \delta_{\mu,-\nu} \mu \sigma_j^z,
  &&
  \sp{S_+,S_-}_- = 2 S_z.
  \label{eq:comm_mu_nu}
\end{align}
\begin{align}
  S_z S_\mu^n
  = \p{\mu S_\mu + S_\mu S_z} S_\mu^{n-1}
  = \mu S_\mu^n + S_\mu S_z S_\mu^{n-1}
  = 2\mu S_\mu^n + S_\mu^2 S_z S_\mu^{n-2}
  = S_\mu^n \p{\mu n + S_z},
  \label{eq:comm_z_mu_n}
\end{align}
\begin{align}
  \sigma_j^+ \sigma_j^- = \f12\p{\1_j+\sigma_j^z},
  &&
  \sum_j \sigma_j^+ \sigma_j^- = S + S_z.
\end{align}
Here $S\equiv N/2$.

@@

The decoherence operator for uncorrelated single-spin decay at a
uniform rate is
\begin{align}
  \D\p{\O}
  \equiv \sum_j\p{\sigma_j^+\O\sigma_j^-
    - \f12\sp{\sigma_j^+\sigma_j^-,\O}_+}
  = \sum_j \sigma_j^+\O\sigma_j^- - S \O - \f12 \sp{S_z,\O}_+.
\end{align}
We are interested in the effect of this decoherence on the dynamics of
collective spin operators, all of which are determined by operators of
the form $S_+^\ell S_-^m S_z^n$.  We thus expand
\begin{align}
  \D\p{S_+^\ell S_-^m S_z^n}
  = \sum_j \sigma_j^+ S_+^\ell S_-^m S_z^n \sigma_j^-
  - S S_+^\ell S_-^m S_z^n - \f12 \sp{S_z, S_+^\ell S_-^m S_z^n}_+.
  \label{eq:D_z+-}
\end{align}
Considering the first term, we can use a commutation relations from
\eqref{eq:comm_z_mu} to push through and simplify the jump operators
as
\begin{align}
  \sum_j \sigma_j^+ S_+^\ell S_-^m S_z^n \sigma_j^-
  = \sum_j S_+^\ell \sigma_j^+ \sigma_j^- S_-^m \p{S_z-1}^n
  = S_+^\ell \p{S + S_z} S_-^m
  \sum_{k=0}^n { n \choose k } \p{-1}^{n-k} S_z^k.
\end{align}
We now need to push the middle $S_z$ operator to the right, which we
can do using the commutation relation $\sp{S_z,S_\mu^m}=\mu mS_\mu$
derived in \eqref{eq:comm_z_mu_n}:
\begin{align}
  \sum_j \sigma_j^+ S_+^\ell S_-^m S_z^n \sigma_j^-
  = S_+^\ell S_-^m \p{S - m + S_z}
  \sum_{k=0}^n { n \choose k } \p{-1}^{n-k} S_z^k.
\end{align}
Considering now the last term in \eqref{eq:D_z+-}, we can expand
\begin{align}
  \f12 \sp{S_z, S_+^\ell S_-^m S_z^n}_+
  = \f12 \p{S_z S_+^\ell S_-^m + S_+^\ell S_-^m S_z} S_z^n,
\end{align}
where we can again use commutation relation in \eqref{eq:comm_z_mu_n}
to push $S_z$ to the right and get
\begin{align}
  S_z S_+^\ell S_-^m
  = S_+^\ell \p{\ell + S_z} S_-^m
  = \ell S_+^\ell S_-^m + S_+^\ell S_z S_-^m
  = S_+^\ell S_-^m \p{\ell - m + S_z},
\end{align}
which implies
\begin{align}
  \f12 \sp{S_z, S_z^\ell S_+^m S_-^n}_+
  = \f12 S_+^\ell S_-^m \p{\ell - m + 2 S_z} S_z^n
  = S_+^\ell S_-^m \p{\f12\sp{\ell-m} + S_z} S_z^n.
\end{align}
Putting everything together, we have that
\begin{align}
  \D\p{S_+^\ell S_-^m S_z^n}
  &= S_+^\ell S_-^m \sp{\p{S - m + S_z} \sum_{k=0}^n
    { n \choose k } \p{-1}^{n-k} S_z^k
    - \p{S + \f12\sp{\ell-m} + S_z} S_z^n} \\
  &= S_+^\ell S_-^m \sp{\p{S - m + S_z} \sum_{k=0}^{n-1}
    { n \choose k } \p{-1}^{n-k} S_z^k - \f12 \p{\ell+m} S_z^n}.
\end{align}

>>>>

Previous work has shown how to realize various spin-squeezing
Hamiltonians with ultracold atomic systems.  While it is
straightforward to simulate the dynamics generated by these
Hamiltonians in the context of a closed quantum system, here we try to
develop an understanding of their the squeezing behavior in the
presence of decoherence.  In particular, we will be considering
dynamics induced by the collective one-axis twisting (OAT), two-axis
twisting (TAT), and transverse field (TVF) Hamiltonians of the form
\begin{align}
  H_{\t{OAT}} = \chi S_z^2,
  &&
  H_{\t{TAT}} = \chi \p{S_z^2 - S_y^2},
  &&
  H_{\t{TVF}} = \chi S_z^2 - \Omega S_x,
\end{align}
for spins which decay in an uncorrelated manner at a rate $\gamma$.
The evolution of any Heisenburg operator $\O$ with a Hamiltonian $H$
is given by
\begin{align}
  \f{d}{dt} \O
  = i\sp{H,\O}_- + \gamma\D\p{\O},
  &&
  \D\p{\O} \equiv \sum_j\p{\sigma_j^+\O\sigma_j^-
    - \f12\sp{\sigma_j^+\sigma_j^-,\O}_+},
  \label{eq:EOM}
\end{align}
where $\sp{A,B}_\pm\equiv AB\pm BA$ and $j$ indexes an individual
spin.  Our measure of squeezing is
\begin{align}
  \xi^2 \equiv \f{N}{\abs{\bk{\v S}}^2}
  \min_{\uv n\perp\bk{\v S}} \bk{\p{\v S\c\uv n}^2},
  \label{eq:squeezing}
\end{align}
where $N$ is the total number of spins, $\v S$ is the collective spin
vector, and the minimization is performed over all unit vectors
$\uv n$ in the plane orthogonal to the mean spin vector $\bk{\v S}$.
This squeezing parameter is entirely determined by components
$\bk{S_\alpha}$ of the mean spin vector in addition to collective
spin-spin correlators of the form $\bk{S_\alpha S_\beta}$; our task is
therefore to compute these expectation values in all scenarios of
interest.

@@

This text frequently makes use of various spin operator identities; we
collect some particularly useful identities here for reference.
\begin{align}
  \f12 \sp{\sigma_j^z,\sigma_k^\mu}_- = \delta_{jk} \mu \sigma_j^\mu,
  &&
  \sp{S_z,\sigma_j^\mu}_- = \f12 \sp{\sigma_j^z,S_\mu}_-
  = \mu \sigma_j^\mu,
  &&
  \sp{S_z,S_\mu}_- = \mu S_\mu
  \label{eq:comm_z_mu}
\end{align}
\begin{align}
  \sp{\sigma_j^\mu,\sigma_k^\nu}_-
  = \delta_{jk} \delta_{\mu,-\nu} \mu \sigma_j^z,
  &&
  \sp{S_\mu,\sigma_j^\nu}_- = \sp{\sigma_j^\mu,S_\nu}_-
  = \delta_{\mu,-\nu} \mu \sigma_j^z,
  &&
  \sp{S_\mu,S_\nu}_- = 2\delta_{\mu,-\nu} \mu S_z
  \label{eq:comm_mu_nu}
\end{align}
\begin{align}
  S_z S_\mu^n
  = \p{\mu S_\mu + S_\mu S_z} S_\mu^{n-1}
  = \mu S_\mu^n + S_\mu S_z S_\mu^{n-1}
  = 2\mu S_\mu^n + S_\mu^2 S_z S_\mu^{n-2}
  = S_\mu^n \p{\mu n + S_z}
  \label{eq:comm_z_mu_n}
\end{align}
\begin{align}
  \sigma_j^+ \sigma_j^- = \f12\p{\1_j+\sigma_j^z}
  &&
  \sum_j \sigma_j^+ \sigma_j^- = S + S_z.
\end{align}
Here $S\equiv N/2$.

@@

The decoherence operator for uncorrelated single-spin decay at a
uniform rate is
\begin{align}
  \D\p{\O}
  \equiv \sum_j\p{\sigma_j^+\O\sigma_j^-
    - \f12\sp{\sigma_j^+\sigma_j^-,\O}_+}
  = \sum_j \sigma_j^+\O\sigma_j^- - S \O - \f12 \sp{S_z,\O}_+.
\end{align}
We are interested in the effect of this decoherence on the dynamics of
collective spin operators, all of which are determined by operators of
the form $S_+^\ell S_-^m S_z^n$.  We thus expand
\begin{align}
  \D\p{S_+^\ell S_-^m S_z^n}
  = \sum_j \sigma_j^+ S_+^\ell S_-^m S_z^n \sigma_j^-
  - S S_+^\ell S_-^m S_z^n - \f12 \sp{S_z, S_+^\ell S_-^m S_z^n}_+.
  \label{eq:D_z+-}
\end{align}
Considering the first term, we can use a commutation relations from
\eqref{eq:comm_z_mu} to push through and simplify the jump operators
as
\begin{align}
  \sum_j \sigma_j^+ S_+^\ell S_-^m S_z^n \sigma_j^-
  = \sum_j S_+^\ell \sigma_j^+ \sigma_j^- S_-^m \p{S_z-1}^n
  = S_+^\ell \p{S + S_z} S_-^m
  \sum_{k=0}^n { n \choose k } \p{-1}^{n-k} S_z^k.
\end{align}
We now need to push the $S_-^m$ to the left, which we can do using the
commutation relation $\sp{S_z,S_\mu^m}=\mu mS_\mu$ derived in
\eqref{eq:comm_z_mu_n}:
\begin{align}
  \sum_j \sigma_j^+ S_+^\ell S_-^m S_z^n \sigma_j^-
  = S_+^\ell S_-^m \p{S - m + S_z}
  \sum_{k=0}^n { n \choose k } \p{-1}^{n-k} S_z^k.
\end{align}
Considering now the last term in \eqref{eq:D_z+-}, we can expand
\begin{align}
  \f12 \sp{S_z, S_+^\ell S_-^m S_z^n}_+
  = \f12 \p{S_z S_+^\ell S_-^m + S_+^\ell S_-^m S_z} S_z^n,
\end{align}
where we can again use commutation relation
$\sp{S_z,S_\mu^m}=\mu mS_\mu$ to push $S_z$ to the right and get
\begin{align}
  S_z S_+^\ell S_-^m
  = S_+^\ell \p{\ell + S_z} S_-^m
  = S_+^\ell S_-^m \p{\ell - m + S_z},
\end{align}
which implies
\begin{align}
  \f12 \sp{S_z, S_+^\ell S_-^m S_z^n}_+
  = \f12 S_+^\ell S_-^m \p{\ell - m + 2 S_z} S_z^n
  = S_+^\ell S_-^m \p{\f12\sp{\ell-m} + S_z} S_z^n.
\end{align}
Putting everything together, we have that
\begin{align}
  \D\p{S_+^\ell S_-^m S_z^n}
  &= S_+^\ell S_-^m \sp{\p{S - m + S_z} \sum_{k=0}^n
    { n \choose k } \p{-1}^{n-k} S_z^k
    - \p{S + \f12\sp{\ell-m} + S_z} S_z^n} \\
  &= S_+^\ell S_-^m \sp{\p{S - m + S_z} \sum_{k=0}^{n-1}
    { n \choose k } \p{-1}^{n-k} S_z^k - \f12 \p{\ell+m} S_z^n}.
\end{align}
